Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-12 infercnv/reports/gains_and_losses.Rmd
add hclust ceiling to pheatmap

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
library(Matrix)
library(SummarizedExperiment)
library(SingleCellExperiment)

@@

}, file = 'sce.rds')
```

@@

```{r plot_umaps_by_id, fig.height = 30}
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2, 
                       split.by = 'sample') +
  coord_fixed()
```

Then, we visualise chr7 and chr5 gains.

```{r plot_chr_gains}
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr7', size = 0.2) +
  coord_fixed()
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr7', size = 0.2, split.by = 'sample') +
  coord_fixed()
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr5', size = 0.2) +
  coord_fixed()
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr5', size = 0.2, split.by = 'sample') +
  coord_fixed()
```

@@

```{r plot_gains_and_losses, fig.height = 20}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::group_by(sample, cell_type) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup()

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::set_names() %>%
  purrr::map(function(direction) {
    p_dat_i <-
      p_dat %>% 
        dplyr::filter(!is.na(paste0('proportion_', direction, '_chr1')))
    list(
      matrix = 
        p_dat_i %>%
        dplyr::select(dplyr::contains(paste0('proportion_', direction))) %>%
        as.matrix(),
      annot = 
        p_dat_i %>% 
        dplyr::select(sample, cell_type)
    )
  })

# plot heatmaps
pheatmap::pheatmap(p_dat$dupli$matrix, cluster_cols = F, cluster_rows = F)
pheatmap::pheatmap(sce@colData[,grepl('proportion_dupli', colnames(sce@colData))])
```

@@

```{r chr7_prop_dist}
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::filter(!is.na(proportion_dupli_chr7))

plot_prop_chr7_gain <- function(p_dat, position_arg) {
  p_dat %>%
    ggplot() +
    geom_bar(aes(x = cell_type, 
                 fill = proportion_dupli_chr7, 
                 group = proportion_dupli_chr7),
             position = position_arg) +
    scale_x_discrete(guide = guide_axis(angle = -45)) +
    theme_classic() +
    scale_fill_viridis_b(n.breaks = 9) 
}

plot_prop_chr7_gain(p_dat, position_stack(reverse = T)) +
  ggtitle('chr7 gain count - all cells')
plot_prop_chr7_gain(p_dat, position_fill(reverse = T)) +
  ggtitle('chr7 gain proportion - all cells')
p_dat %>%
  {split(., .$id)} %>%
  purrr::map2(names(.), ., function(id, df) {
    plot_prop_chr7_gain(df, position_fill(reverse = T)) +
      ggtitle(paste0('chr7 gain count - ', id))
    plot_prop_chr7_gain(df, position_stack(reverse = T)) +
      ggtitle(paste0('chr7 gain proportion - ', id))
  })
```

@@

We pseudobulk chr7+ cells and perform differential expression analysis. In order to do this, we use the `glmGamPoi` package in R. This analysis follows [this tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html).

@@

  # pseudobulk the sce object by patient, celltype, and chr7 ploidy
  reduced_sce <-
    glmGamPoi::pseudobulk_sce(
      sce, 
      group_by = vars(sample, cell_type, chr7_gain),
      n_cells = n()
    )

  # remove NAs
  reduced_sce <- reduced_sce[, !is.na(reduced_sce$cell_type)]

@@

}, file = 'reduced_sce.rds')
```

@@

```{r deseq, message = T}
# use DESeq2's size factor calculation procedure
fit <-
  glmGamPoi::glm_gp(
    reduced_sce,
    design = ~ chr7_gain * cell_type + sample,
    size_factor = 'ratio',
    verbose = T
  )
res <-
  glmGamPoi::test_de(
    fit, 
    contrast = 
      cond(cell_type = 'Stromal cell', chr7_gain = 1) -
      cond(cell_type = 'Stromal cell', chr7_gain = 0)
  )
```

```{r plot_de}
res %>%
  ggplot(aes(x = lfc, y = log10(pval))) +
  geom_point(aes(colour = adj_pval < 0.01), size = 0.5) +
  t
```

>>>>

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
library(Matrix)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(glmGamPoi)

@@

}, file = 'sce.rds', rerun = params$rerun)
```

@@

Then, we visualise chr7 gains on the UMAP.

```{r plot_chr_gains}
dittoSeq::dittoDimPlot(
  sce, 'proportion_dupli_chr7', size = 0.1) +
  coord_fixed()
```

@@

First, we plot single cell-level CNVs by direction and by celltype.

```{r plot_gains_and_losses_by_ct, fig.height = 20, eval = F}
# at single-cell level
unique(sce$cell_type) %>%
  purrr::walk(function(ct) {
    c('dupli', 'loss') %>%
      purrr::walk(function(direction) {
        p_dat <-
          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
        p_mat <-
          p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
        p_annot <- 
          p_dat[, c('sample'), drop = F] %>%
          as.data.frame()
        p <-
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
            legend = F,
            main = paste0('celltype: ', ct, '\ndirection: ', direction))
        print(p)
      })
  })
```

We also plot sample-x-celltype level events (mean-summarised).

```{r plot_gains_and_losses_by_id}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::group_by(sample, cell_type) %>%
  dplyr::filter(!is.na(proportion_cnv_chr1)) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(sample_x_ct = paste0(sample, '_', cell_type)) %>%
  tibble::column_to_rownames('sample_x_ct')

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c('sample', 'cell_type')] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0('direction: ', direction))
  })
```

@@

```{r chr7_prop_dist}
# prep data
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::filter(!is.na(proportion_dupli_chr7)) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(n_cells = dplyr::n())

# define functions
plot_prop_chr7_gain <- function(p_dat, title) {
  plot_func <- function(p_dat, position_arg) {
    p_dat %>%
      ggplot() +
      geom_bar(aes(x = cell_type, 
                   fill = proportion_dupli_chr7, 
                   group = proportion_dupli_chr7),
               position = position_arg) +
      scale_x_discrete(guide = guide_axis(angle = -45)) +
      theme_classic() +
      scale_fill_viridis_b(n.breaks = 9, limits = c(0, 1)) 
  }

  # plot both count (stack) and proportion (fill), annotate n cells
  list(
    plot_func(p_dat, position_fill(reverse = T)) +
      ggtitle(title) +
      theme(axis.text.x = element_blank(),
            axis.title.x = element_blank()) +
      ylab('count'),
    plot_func(p_dat, position_stack(reverse = T)) +
      ylab('proportion')) %>%
    patchwork::wrap_plots(ncol = 1) 
}

# plot, across all and by patient
plot_prop_chr7_gain(p_dat, 'chr7 gain - all cells')
p_dat %>%
  {split(., .$sample)} %>%
  purrr::walk2(names(.), ., function(id, df) {
    df %>%
    plot_prop_chr7_gain(
      paste0('chr7 gains - ', id, ' (', unique(df$n_cells), ' cells)')) %>%
      print()
  })
```

@@

We pseudobulk chr7+ cells and perform differential expression analysis. In order to do this, we use the `glmGamPoi` package in R. This analysis follows a [tutorial](https://bioconductor.org/packages/devel/bioc/vignettes/glmGamPoi/inst/doc/pseudobulk.html) and the [glmGamPoi GitHub](https://github.com/const-ae/glmGamPoi).

@@

  # pseudobulk the sce object by patient and chr7 ploidy
  # subset to stromal cells
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, sce$cell_type == 'Stromal cell'], 
      group_by = vars(sample, chr7_gain)
    )

  # remove NA's
  reduced_sce <-
    reduced_sce[, !is.na(reduced_sce$sample) & !is.na(reduced_sce$chr7_gain)]

@@

}, file = 'reduced_sce.rds', rerun = params$rerun)
```

@@

We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status and the patient as factors.

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
  glmGamPoi::glm_gp(
    reduced_sce,
    design = ~ chr7_gain + sample,
    verbose = T
  )
  fit
}, file = 'fit.rds', rerun = params$rerun)

# show model output
summary(fit)
```

Next, we test for differential gene expression.

```{r test_ge}
# get differential expression
de_res <-
  glmGamPoi::test_de(
    fit, 
    contrast = 
      cond(chr7_gain = 1) -
      cond(chr7_gain = 0)
  )
```

```{r plot_de}
# minimum accepted adj_pval
min_p <- 0.05

# define up/downregulation
de_res <-
  de_res %>% 
  tibble::as_tibble() %>%
  dplyr::mutate(
    direction = dplyr::case_when(
      lfc < -1.5 & adj_pval < min_p ~ 'downregulated',
      lfc > 1.5 & adj_pval < min_p ~ 'upregulated',
      TRUE ~ NA
    )
  )

# plot
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name),
    min.segment.length = 0, nudge_x = 0.5) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey')
```

These are the most strongly downregulated genes:

```{r get_downreg_degs}
de_res %>%
  tibble::as_tibble() %>% 
  dplyr::filter(adj_pval < 0.01) %>%
  dplyr::arrange(-lfc, adj_pval) %>%
  head(10) %>%
  knitr::kable()
```

These are the most strongly upregulated genes:

```{r get_upreg_degs}
de_res %>%
  tibble::as_tibble() %>% 
  dplyr::filter(adj_pval < 0.01) %>%
  dplyr::arrange(lfc, adj_pval) %>%
  head(10) %>%
  knitr::kable()
```
